---
title: "DESeq2 post analysis plots of expression values with a multipage PDF catalog"
author: "Stephen Turner"
date: "June 6, 2016"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE)
```

## The point

I get asked all the time, "hey, could you make a boxplot of the expression values of _SomeGene_ for me?" The `plotCounts()` function in DESeq will do this, but only one gene at a time. This shows you how to do this for any arbitrary number of genes in your data (or all of them!). You can plot a small number in a single plot with facets, or you could plot a large number in a text-searchable multi-page PDF.

## Setup

Load the packages you'll use.

```{r}
library(tidyverse)
library(DESeq2)
library(airway)
theme_set(theme_bw(base_size=14) + theme(strip.background = element_blank()))
```

Load the data from the airway package, and add a made up "group" variable. We'll use this downstream when we map aesthetics onto a boxplot. Create a `DESeqDataSet` and run the DESeq pipeline on it.

```{r}
# Load the data
data(airway)

# Make up a fake "group" variable
colData(airway)$fake <- factor(rep(c("groupA", "groupB"), each=4))

# Make a dataset and run the pipeline
dds <- DESeqDataSet(airway, design=~cell+dex)
dds <- DESeq(dds)
```

Inspect the `colData`.

```{r}
colData(dds)
```

## Get results

Get the results for the dex treated vs untreated samples and arrange by p-value.

```{r}
res <- results(dds, tidy=TRUE, contrast=c("dex", "trt", "untrt")) %>%
  arrange(padj, pvalue) %>%
  tbl_df()
res
```

Now, define some genes of interest. These must be the names of the genes as in the count data. That is, your genes of interest must be `%in% names(dds)`. These are the genes you'll plot. You could do as many as you like. Since you've arranged by p-value, let's take the top few results (9 in this case, so we make a square faceted plot).

```{r}
# Define the genes of interest.
goi <- res$row[1:9]
stopifnot(all(goi %in% names(dds)))
goi
```

## Join & tidy

Create a tidy/transposed count matrix. Here's a line-by-line explanation: 

1. Here, you'll take the `counts()` of the dds object, normalizing it, without outlier replacement. You'll add a half count to it, because the next thing you'll do is `log2()` it, and you don't want any `-Inf`s. So now you have a log-transformed normalized count matrix. Now, transpose that matrix so the sample names are the `row.names`. 
2. You'll now merge that thing with the `colData(dds)`, where the sample names are also the `row.names`. You'll now have a really wide data.frame with a single row for each sample, followed by all the colData, followed by a column for each of the genes of interest.
3. Gather that up again so you now have one row per sample per gene, with all the sample info, and that gene's expression.

```{r}
tcounts <- t(log2((counts(dds[goi, ], normalized=TRUE, replaced=FALSE)+.5))) %>%
  merge(colData(dds), ., by="row.names") %>%
  gather(gene, expression, (ncol(.)-length(goi)+1):ncol(.))
```

Take a look. 

```{r}
tcounts %>% 
  select(Row.names, dex, fake, gene, expression) %>% 
  head %>% 
  knitr::kable()
```

## Single faceted plot

Now, create a single faceted plot. You don't have to add a fill aesthetic - I only did so as an example, using the fake variable I made up.

```{r, fig.height=8, fig.width=8}
ggplot(tcounts, aes(dex, expression, fill=fake)) + 
  geom_boxplot() + 
  facet_wrap(~gene, scales="free_y") + 
  labs(x="Dexamethasone treatment", 
       y="Expression (log normalized counts)", 
       fill="(Some made \nup variable)", 
       title="Top Results")
```

## Multi-page "catalog"

Or in a loop, create a multi-page PDF (text searchable) with one page per gene.

```{r, eval=FALSE}
pdf("multi-ggplot2-catalog.pdf")
for (i in goi) {
  p <- ggplot(filter(tcounts, gene==i), aes(dex, expression, fill=fake)) + 
    geom_boxplot() + 
    ggtitle(i)
  print(p)
}
dev.off()
```

## Improvements

- In the "Join & tidy" section, you could continue joining back to the results, such that you could incorporate statistics (FCs, p-values, etc) into the text on the plots. 
- You could also consider joining this to gene information, e.g., using [annotables](https://github.com/stephenturner/annotables) or Bioconductor annotation packages.
