---
title: "Playing around with networkD3"
author: "Stephen Turner"
date: "3/24/2017"
output: 
  html_document: 
    toc: no
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Simple stuff

```{r}
library(networkD3)
```

Create some fake data.

```{r}
networkData <- data.frame(
  source = c("A", "A", "A", "A", "B", "B", "C", "C", "D"), 
  target = c("B", "C", "D", "J", "E", "F", "G", "H", "I")
)
networkData
```

Plot it. Try hovering, zooming, clicking/dragging nodes, etc.

```{r}
simpleNetwork(networkData, zoom=TRUE, fontSize = 10)
```

## But wait Jean Valjean, There's More

Let's use some linkages and metadata about characters from _Les Miserables_.

```{r}
data(MisNodes)
head(MisNodes, 15)

data(MisLinks)
head(MisLinks, 10)
```

How's about a [force-directed graph drawing](https://en.wikipedia.org/wiki/Force-directed_graph_drawing)? Hover over some of the points, use the mouse wheel to zoom in/out, click and drag around, etc. I'll need to learn a bit more about the data structures going into this...

```{r}
forceNetwork(Links = MisLinks, Nodes = MisNodes,
            Source = "source", Target = "target",
            Value = "value", NodeID = "name", Group = "group", 
            opacity = 0.8, fontSize=10, zoom=TRUE)
```

