---
title: "Forecasting"
author: "Stephen Turner"
date: "November 16, 2017"
output: 
  html_document: 
    code_folding: hide
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    theme: flatly
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
# subtitle: preprocessing and analysis
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
```

## Forecasting flu outbreaks

```{r}
library(tidyverse)
library(prophet)
library(cdcfluview)
library(lubridate)
theme_set(theme_bw())
```

### Get data

Let's get the data from the CDC's Influenza-like illness API. 

```{r}
flu <- ilinet(region = "national") %>% 
  select(week_start, starts_with("age"), ilitotal, total_patients) %>%
  filter(year(week_start)>=2003)
flu
```

### Forecast future outbreaks

The last week we have data on is `r tail(flu$week_start,1)`. Let's forecast outbreaks for the next five years.

```{r, results="hide"}
# Run the model
m <- prophet(flu %>% select(ds=week_start, y=ilitotal))
# Make a future data frame
future <- make_future_dataframe(m, periods=365*5)
# Predict the future
forecast <- predict(m, future)
tail(forecast[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])
```

### Plot the future

```{r, echo=FALSE, fig.cap="Flu outbreaks forecasted for the next 5 years."}
plot(m, forecast) + labs(x="Date", y="Number of flu cases")
```

```{r, echo=FALSE, fig.cap="Trend, uncertainty, and yearly seasonality."}
prophet_plot_components(m, forecast)
```

## Forecast Peyton Manning Wikipedia Page Views

Let's look at a time series of daily page views for the Wikipedia page for [Peyton Manning](https://en.wikipedia.org/wiki/Peyton_Manning) scraped using the [Wikipediatrend package](https://cran.r-project.org/web/packages/wikipediatrend/vignettes/using-wikipediatrend.html). This example illustrates forecasting that accounts for multiple seasonality, changing growth rates, and the ability to model special days (such as Manning's playoff and superbowl appearances). The data is available [here](https://raw.githubusercontent.com/stephenturner/sandbox/master/data/peyton_manning_prophet.csv).

### Process data and forecast

Import and log transform the data to deal with extreme outliers, then predict the future.

```{r, results="hide"}
manning <- read_csv("data/peyton_manning_prophet.csv") %>% mutate(y=log(y))
m <- prophet(manning)
future <- make_future_dataframe(m, periods = 365*2)
forecast <- predict(m, future)
tail(forecast[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])
```

### Plot future page views

```{r, echo=FALSE, fig.cap="Predicted page views for Wikipedia entry on Peyton Manning."}
plot(m, forecast) + labs(x="Date", y="Number of page views")
```

```{r, echo=FALSE, fig.cap="Trend/uncertainty, and weekly + yearly seasonality."}
prophet_plot_components(m, forecast)
```
