---
title: "Tidyeval / rlang playground"
author: "Stephen Turner"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    # code_folding: show
    # df_print: paged
    out.width: "100%"
    fig.width: 9
    fig.height: 3
    warning: FALSE
    message: FALSE
    fig_caption: yes
    theme: flatly
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
# subtitle: asdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
```

# Useful references

Official documentation: 

- rlang documentation: http://rlang.tidyverse.org/
- Programming with dplyr: http://dplyr.tidyverse.org/articles/programming.html

Useful posts:

- http://www.onceupondata.com/2017/08/12/my-first-steps-into-the-world-of-tidyeval/
- http://blog.jalsalam.com/posts/2017/quasi-quotation-applications/

# Recoding with a lookup table

```{r}
d <- data_frame(x=1:5, y=letters[1:5])
d
```

Let's recode one row at a time

```{r}
d %>% 
  mutate(z=recode(y, a="apple", b="banana", c="cat", d="dog", e="elephant")) # etc
```

What if we have a lookup dictionary in a table? You might read this in from a file, but here we create it. Notice that you don't have an entry for `d`, but you do for `f`. Notice the effect later on.

```{r}
lookup <- tribble(
  ~letter, ~expansion, 
  "a", "apple",
  "b", "banana", 
  "c", "cat",
  "e", "elephant",
  "f", "fox"
)
lookup
```

Make it a named list

```{r}
lookup <- deframe(lookup)
lookup
```

Unquote splice (`!!!`) to recode all.

```{r}
d %>% 
  mutate(z=recode(y, !!!(lookup)))
```

Notice what happens for `d`. Use this to make it default back to missing.

```{r}
d %>% 
  mutate(z=recode(y, !!!(lookup), .default=NA_character_))
```
